---
title: "Project proposal"
author: "Fantastic 4"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
df <- read_csv("../data/vdem3.csv")
```

## 1. Introduction

This project deals with economy, democracy, and dictatorship. Specifically, we inquire on the relationship between countries’ economy and their regime. We ask how a country’s economy determines its regime measured in democracy and dictatorship continuum. We further ask specific questions:

(1) How does the region mediate the relationship? 
(2) What historically relevant events one notices that affect both the economy and regime?
(3) Can it be that the regime determines the economy?

To answer these questions, we use regime and economic indicators collected from the Varieties of Democracy (V-Dem) dataset and World Development Indicators (WDI) dataset of the World Bank, respectively. The V-Dem dataset provides us with five indices of democracy: polyarchy (or electoral), egalitarian, deliberative, liberal, and participatory democracy. 

To supplement this, the WDI dataset provides economic indicators using Gross Domestic Product per capita (GDP per capita). Finally, our working dataset includes 181 countries and their three-letter ID spanning from 1960 to 2024, along with historically relevant events, their date, and region. Thus, each observation is a country’s regime and economic data in a given year. 

## 2. Data

Text goes here. Place your data in the /data folder, and add dimensions and codebook to the README in that folder. Then print out the output of glimpse() or skim() of your data frame.

```{r}
# Code goes here
# Read in your data file
# Print the output of glimpse() or skim()
```

## 3. Data analysis plan

Text goes here.
- What variables will you visualize to explore your research questions?
- Will there be any other data that you need to find to help with your research question?
- Very preliminary exploratory data analysis, including some summary statistics and visualizations, along with some explanation on how they help you learn more about your data. (You can add to these later as you work on your project.)
- The data visualization(s) that you believe will be useful in exploring your question(s). (You can update these later as you work on your project.)


```{r}
# Code goes here
# Code to calculate summary statistics
# Code for a visualization
```
